---
title: 'Data cleaning: missing data'
author: "Zena Lapp"
date: "2023-01-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Today we're going to learn how to clean up a dataset with missing data. 

### Read in and look at data

First, we need to read in our data and take a look at it. It's in the `data` folder.

```{r}
library(tidyverse)
dat <- read_csv('data/dat.csv')
# downloads folder:
dat <- read_csv('~/Downloads/missing_data_code_club/data/dat.csv')
# desktop folder:
dat <- read_csv('~/Desktop/missing_data_code_club/data/dat.csv')

# look at data
dat
```

There are a few things going on:

1. Sex is not consistently categorized. 
1. There are missing values (coded as `NA` in R).
1. Age is a character, but we'd expect it to be a double (number). 
1. The units seem off (age is in months, height is in inches, and weight is in pounds).

Today we will be focusing on cleaning up the missing data. In subsequent code clubs, we will clean up other parts of the dataset. 

### Looking into the age column

First, let's take a look at what's going on with the age column by seeing which values aren't numbers.

```{r}
dat %>%
    filter(is.na(as.numeric(age)))
```

It seems like some of the missing values in age are coded as 'N/A' and one column has "months" written in as well. Let's change the N/A's to `NA`, which is recognized by R as a missing value. 

```{r}
dat_clean_age <- dat %>%
    mutate(age = replace(age, age == 'N/A', NA))
```

Did the code do what we wanted it to?

```{r}
dat_clean_age %>%
    filter(age == 'N/A')
dat %>%
    filter(age == 'N/A')
```

Great! 

### Count number of missing values in each column

Now, how many missing values to we have in the sex and age columns?

```{r}
dat_clean_age %>%
    summarize(sex_na = sum(is.na(sex)),
              age_na = sum(is.na(age)))

dat_clean_age %>%
    mutate(sex_na = is.na(sex),
           sex_na = as.numeric(sex_na))

# all columns together
dat_clean_age %>%
    summarize_all(funs(sum(is.na(.))))
```

Add to the code above to count the NA's for sex, height, and weight.

```{r}

```


### Dealing with missing values

Okay, now we have a sense of how many NAs there are in our dataset. How do we deal with them? 
1. We can remove rows with missing data from the dataset.
1. We can impute missing data from the other columns. We won't have time to do this now, but you can learn more about it [here](https://www.simplilearn.com/data-imputation-article). 

Remove rows with missing data:

```{r}
dat_no_missing <- dat_clean_age %>%
    drop_na()
```

Check for missing data:

```{r}
dat_no_missing %>%
    summarize_all(funs(sum(is.na(.))))
```

How many rows were dropped?

```{r}
nrow(dat_clean_age) - nrow(dat_no_missing)
```


## Bonus: Cleaning up the age column even more

Starting with `dat_clean_age`, change "174 months" to "174" and then change the age column to numeric.

```{r}
dat_clean_age %>%
    mutate(age = replace(age, age == '174 months', 174),
           age = as.numeric(age))

# a little more generalized
dat_clean_age %>%
    filter(age == '174 months') %>% 
    mutate(age = gsub('months', '', age),
           age = as.numeric(age))

gsub('hi*', 'hello', 'hiiiiiii my name is zena')
```

