---
title: "Calculating percentages"
author: "Code club"
date: '2022-09-27'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Finding percentages using `dplyr` can be a little bit complicated. However, it’s a very useful skill! Today, we're going to figure out how to do it. 

We are going to use a built-in dataset in the tidyverse called `population` that gives us the population of each country across different years.

First, we have to load the tidyverse:

```{r}

```

Next, let's take a look at the dataset:

```{r}

```

### Goal 1: Calculate the percentage of the global population that each country makes up for the year 2000.

First, think about the exact steps you need to take to accomplish your goal, and write these down. This is called writing "pseudocode", which is a great step to take before actually writing your code because it helps you figure out the logic of what you need to do without having to worry about the syntax. 

Once you've written out your pseudocode, try coding it up. Save your output to an object called `pop_pcts_2000`.

```{r}

```

Using the data you just created, figure out the answers to the questions below.

What percentage of the global population does Kenya make up for the year 2000? 

```{r}

```

Which country makes up the highest percent of the population? The lowest?

```{r}

```

### Goal 2: Calculate the percentage of the global population that each country makes up for each year in the dataset. 

Again, write pseudocode first to help you figure out exactly what you need to do. Once you've done this, write up the code and save the output to an object called `pop_pcts`.

```{r}

```

With this new dataset, determine whether the population for Kenya increasing or decreasing over time.

```{r}

```


