---
title: "Cleaning messy columns"
author: "Zena Lapp"
date: "2023-02-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Today we're going to go over how to recode information in columns and to create new columns. 

First, run this code chunk to load tidyverse and generate the data we're going to use:

```{r}
# load library
library(tidyverse)

# set seed so we all have the same data
set.seed(1)

# generate dataset
messy_data <- tibble(id = 1:100,
                     gender = sample(c('F', 'Female', 'female', 'M', 'Male', 'male'), 
                                     size = 100, replace = TRUE),
                     age_months = round(rnorm(100, mean = 30*12, sd = 10*12))) 
```

Let's take a look at our dataset: 

```{r}
messy_data 
```

It appears that gender is not coded consistently. Let's take a look at the different ways it's coded:

```{r}
messy_data %>% 
  group_by(gender) %>% 
  count()
```

Let's clean this data up. There are lots of ways to do this. We'll go over one way.

```{r}
clean_data <- messy_data %>% 
  mutate(gender = case_when(gender %in% c('F', 'Female', 'female') ~ 'female',
                            gender %in% c('M', 'Male', 'male') ~ 'male'))
```

Did our code do what we wanted it to?

```{r}
clean_data %>% 
  group_by(gender) %>% 
  count()
```
Looks good! Now let's make an age in years column.

```{r}
all_data <- clean_data %>% 
  mutate(age_years = age_months/12) %>% 
  select(-age_months)
all_data %>% 
  arrange(age_years)
```

**Bonus:**

make a new column that categorizes age in years as <20, 20-40, >40

```{r}
all_data %>% 
  mutate(age_cat = case_when(age_years < 20 ~ '<20',
                             age_years >= 20 & age_years <= 40 ~ '20-40',
                             age_years > 40 ~ '>40'))
```

How many of each age category do we have?

```{r}
all_data %>% 
  mutate(age_cat = case_when(age_years < 20 ~ '<20',
                             age_years >= 20 & age_years <= 40 ~ '20-40',
                             age_years > 40 ~ '>40')) %>% 
  group_by(age_cat) %>% 
  tally()
```

