---
title: "Logic & Conditionals"
author: "Zena Lapp"
date: "2023-04-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Today we are going to go over logic and conditionals, like we might want to use when filtering rows with `filter()` or when updating or creating a new column with `case_when()`.

First, run this code chunk to generate the data we will be working with:

```{r}
library(tidyverse)

dat <- tibble(person_id = 1:7,
              symptoms = c('yes', 'no', 'yes', 'no', 'no', 'no', 'yes'),
              test_result = c('positive', 'negative', 'positive', 'positive', 'undetermined', 'negative', 'negative'),
              medication = c('yes', 'no', 'no', 'yes', 'no', 'unknown', 'yes'))
```

Let's take a look at the data:

```{r}
dat
```

Some basic logic and how to code it in R:

- Test for equality: double equals (`==`)
- Test for inequality: exclamation mark equals (`!=`)
- Test if something is false: exclamation mark (`!`)
- Test if two things are true: ampersand (`&`)
- Test if at least one thing is true: pipe (`|`)  

```{r}
1 == 1
1 != 1
1 == 2
1 != 2
!(1 == 1)

1 == 1 & 2 == 2
1 == 1 & 2 != 2
1 == 1 | 2 == 2
1 == 1 | 2 != 2

!(1 == 1 | 2 != 2)
1 == 1 | !2 != 2
```

Now, let's see how we might want to use it with the dataset from above.

First, let's filter out the people with missing data. Then we'll answer the below questions.

Questions:

1. Which people had symptoms?
1. Which people tested positive?
1. Which people had symptoms and tested negative?
1. Which people had symptoms and took medication? 
1. Which people did not have symptoms and tested positive? 
1. Which people had symptoms, tested negative, and took medication? 

```{r}
dat %>% 
  filter(test_result != 'undetermined' & medication != 'unknown') %>% 
  mutate(symptoms_yes = symptoms == 'yes',
         test_pos = test_result == 'positive',
         symp_neg = symptoms_yes & !test_pos,
         symp_med = symptoms_yes & medication == 'yes',
         no_symp_pos = !symptoms_yes & test_pos,
         symp_neg_med = symptoms_yes & !test_pos & medication == 'yes')
```

Now, let's classify each person as:

- Symptomatic, asymptomatic, uninfected, or unknown
- Took medication when positive, took medication when negative, or did not take medication

```{r}
dat %>% 
  mutate(case_type = case_when(symptoms == 'yes' & test_result == 'positive' ~ 'symptomatic',
                               symptoms == 'no' & test_result == 'positive' ~ 'asymptomatic',
                               test_result == 'negative' ~ 'uninfected',
                               TRUE ~ 'unknown'),
         med_type = case_when(test_result == 'positive' & medication == 'yes' ~ 'med positive',
                              test_result == 'negative' & medication == 'yes' ~ 'med negative',
                              TRUE ~ 'no med'))
```

